\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-07-22 paper.qmd
Resize Figure 8

mse_scurve_b <- ggplot(error_scurve, 
                     aes(x = b, 
                         y = log(MSE),
                         color = buff,
                         group = buff)) + 
  geom_point(size = 0.05) +
  geom_line(linewidth = 0.05) + 
  geom_vline(xintercept = 220, linetype="solid", 
             color = "black", linewidth=0.1) +
  geom_vline(xintercept = 112, linetype=2, 
             color = "black", linewidth=0.1) +
  geom_vline(xintercept = 364, linetype=2, 
             color = "black", linewidth=0.1) +
  scale_colour_continuous_sequential(palette = "Magenta") +
  labs(x = expression(b), y = "log(MSE)", color = expression(q)) +
  ggtitle("(a)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7, 
            angle=90),
        title = element_text(size = 72),
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 5))
```

@@

```{r}
cell_count_scurve <- ggplot(df_bin_centroids_scurve, 
                            aes(x = reorder(as.factor(hexID), 
                                            -std_counts), 
                                y = std_counts)) + 
  geom_quasirandom(
    size = 0.3,
    alpha = 0.5
  ) + 
  geom_hline(yintercept = benchmark1, color = "black", linewidth=0.5, alpha = 0.5) +
  xlab("hexagon id (re-ordered)") + 
  ylab("standardized bin count") +
  ggtitle("(b)") +
  theme(aspect.ratio = 0.75,
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 5, angle = 90),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 5),
        title = element_text(size = 36))
```

The benchmark value for removing low-density hexagons ranges between $0$ and $1$. When analyzing how these benchmark values influence model performance, it's essential to observe the change in MSE as the benchmark value increases (see @fig-mse-scurve-lwd). The MSE shows a gradual increase as the benchmark value progresses from $0$ to $1$. Evaluating this rate of increase is important. If the increment is not considerable, the decision might lean towards retaining low-density hexagons.

@@

mse_scurve_lwd <- ggplot(error_rm_scurve, 
                     aes(x = benchmark_rm_lwd, 
                         y = log(MSE))) + 
  geom_point(
    size = 0.6
    ) +
  geom_line(
    linewidth = 0.3
    ) + 
  geom_vline(xintercept = benchmark1, linetype="solid", 
             color = "black", linewidth=0.5, alpha = 0.5) +
  scale_x_continuous("standardized bin count", 
         transform = "reverse") + 
  ylab("log(MSE)") +
  ggtitle("(c)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 5),
        axis.text.y = element_text(size = 5),
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 5, 
            angle=90),
        title = element_text(size = 72))
```

Furthermore, selecting the benchmark value for removing low-density hexagons is important. Removing unnecessary bins may lead to the formation of long edges and an uneven \gD{} model. Hence, rather than solely relying on the benchmark value to identify hexagons for removal, it's essential to consider the standard number of points in the neighboring hexagons of the identified low-density bins (see @fig-lwd-scurve (b)). If neighboring bins also show low counts, only those bins will be removed. The remaining bins are used to construct the \gD{} model.

@@

1. Plot the distribution of the 2D Euclidean distances (@fig-param-scurve (d)).
2. Identify the first largest difference between consecutive distance values.
3. Take the distance value corresponding to this difference as the benchmark value.

@@

## To draw the distance distribution
distance_scurve$group <- "1"
distance_scurve_plot <- ggplot(distance_scurve, 
                        aes(x = group, 
                            y = distance)) + 
  geom_quasirandom(
    size = 0.5, 
    alpha = 0.3
  ) + 
  ylim(0, max(unlist(distance_scurve$distance))+ 0.5) + 
  coord_flip() + 
  geom_hline(yintercept = benchmark_scurve,
     linetype="solid", 
     color = "black", linewidth=0.5, alpha = 0.5) +
  ylab(expression(d^{(2)})) +
  ggtitle("(d)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 5),
        axis.title.x = element_text(size = 5),
        title = element_text(size = 36))
```

@@

<!--trimesh with different benchmark values-->

```{r}
## With calculated default: 0.201
trimesh_lg_scurve1 <- vis_rmlg_mesh(
  distance_edges = distance_scurve, 
  benchmark_value = benchmark_scurve, 
  tr_coord_df = tr_from_to_df_scurve, 
  distance_col = "distance") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a", sc_ltr_pos)

trimesh_lg_scurve2 <- vis_rmlg_mesh(
  distance_edges = distance_scurve, 
  benchmark_value = 0.35, 
  tr_coord_df = tr_from_to_df_scurve, 
  distance_col = "distance") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos)

trimesh_lg_scurve3 <- vis_rmlg_mesh(
  distance_edges = distance_scurve, 
  benchmark_value = 1.1, 
  tr_coord_df = tr_from_to_df_scurve, 
  distance_col = "distance") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c", sc_ltr_pos) 
```

```{r}
#| echo: false
#| label: fig-lg-scurve
#| fig-pos: H
#| fig-cap: "$2\\text{-}D$ model generated for different benchmark values to remove long edges: (a) benchmark = $0.201$ (default), (b) benchmark = $0.35$, and (c) benchmark = $1.1$. What is the effective benchmark value to remove long edges?"
#| fig-width: 6
#| fig-height: 3
#| eval: false

trimesh_lg_scurve1 + trimesh_lg_scurve2 + trimesh_lg_scurve3 +
  plot_layout(guides='collect', ncol = 3) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

@@

## A curious difference between t-SNE, UMAP and PaCMAP revealer

@@

The tSNE layout, model in \gD{}, and a view of the fit in projections from $4\text{-}D$, for the five Gaussian cluster data. The model fits the separation and tries to *filled out* the clusters. (The **langevitour** software is used to view the data with a tour, and the full video is available at (<https://youtu.be/RASEE7N5MbM>). 
:::

@@

The UMAP layout, model in \gD{}, and a view of the fit in projections from $4\text{-}D$, for the five Gaussian cluster data. The model fits the separation and tries to *filled out* the clusters, but not as much as tSNE. (The **langevitour** software is used to view the data with a tour, and the full video is available at (<https://youtu.be/iG4bCPkJilw>). 
:::

@@

The PaCMAP layout, model in \gD{}, and a view of the fit in projections from $4\text{-}D$, for the five Gaussian cluster data. The model fits the separation and shows *flat* shaped clusters. (The **langevitour** software is used to view the data with a tour, and the full video is available at (<https://youtu.be/z07cKXi8EJQ>). 
:::

>>>>

mse_scurve_b <- ggplot(error_scurve, 
                     aes(x = b, 
                         y = log(MSE),
                         color = buff,
                         group = buff)) + 
  geom_point(size = 1) +
  geom_line(linewidth = 0.05) + 
  geom_vline(xintercept = 220, linetype="solid", 
             color = "black", linewidth=0.1) +
  geom_vline(xintercept = 112, linetype=2, 
             color = "black", linewidth=0.1) +
  geom_vline(xintercept = 364, linetype=2, 
             color = "black", linewidth=0.1) +
  scale_colour_continuous_sequential(palette = "Magenta") +
  labs(x = expression(b), y = "log(MSE)", color = expression(q)) +
  ggtitle("(a)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12, 
            angle=90),
        plot.title = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        legend.key.height = unit(1, 'cm'),
        legend.key.width = unit(1, 'cm'),
        legend.key.size = unit(1, 'cm'))
```

@@

```{r}
cell_count_scurve <- ggplot(df_bin_centroids_scurve, 
                            aes(x = reorder(as.factor(hexID), 
                                            -std_counts), 
                                y = std_counts)) + 
  geom_quasirandom(
    size = 1,
    alpha = 0.5
  ) + 
  geom_hline(yintercept = benchmark1, color = "black", linewidth=0.5, alpha = 0.5) +
  xlab("hexagon id (re-ordered)") + 
  ylab("standardized bin count") +
  ggtitle("(b)") +
  theme(aspect.ratio = 0.75,
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12, angle = 90),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12),
        plot.title = element_text(size = 12))
```

The benchmark value for removing low-density hexagons ranges between $0$ and $1$. When analyzing how these benchmark values influence model performance, it's essential to observe the change in MSE as the benchmark value increases (@fig-param-scurve). The MSE shows a gradual decrease as the benchmark value goes from $1$ to $0$. Evaluating this rate of increase is important. If the increment is not considerable, the decision might lean towards retaining low-density hexagons.

@@

mse_scurve_lwd <- ggplot(error_rm_scurve, 
                     aes(x = benchmark_rm_lwd, 
                         y = log(MSE))) + 
  geom_point(
    size = 1
    ) +
  geom_line(
    linewidth = 0.3
    ) + 
  geom_vline(xintercept = benchmark1, linetype="solid", 
             color = "black", linewidth=0.5, alpha = 0.5) +
  scale_x_continuous("standardized bin count", 
         transform = "reverse") + 
  ylab("log(MSE)") +
  ggtitle("(c)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12, 
            angle=90),
        plot.title = element_text(size = 12))
```

<!-- Furthermore, selecting the benchmark value for removing low-density hexagons is important. Removing unnecessary bins may lead to the formation of long edges and an uneven \gD{} model. Hence, rather than solely relying on the benchmark value to identify hexagons for removal, it's essential to consider the standard number of points in the neighboring hexagons of the identified low-density bins (see @fig-lwd-scurve (b)). If neighboring bins also show low counts, only those bins will be removed. The remaining bins are used to construct the \gD{} model.    -->

@@

1. Plot the distribution of the \gD{} Euclidean distances (@fig-param-scurve (d)).
2. Identify the first largest difference between consecutive distance values.
3. Take the distance value corresponding to this difference as the benchmark value.

@@

## To draw the distance distribution
distance_scurve$group <- "1"
distance_scurve_plot <- ggplot(distance_scurve, 
                        aes(x = group, 
                            y = distance)) + 
  geom_quasirandom(
    size = 2, 
    alpha = 0.3
  ) + 
  ylim(0, max(unlist(distance_scurve$distance))+ 0.5) + 
  coord_flip() + 
  geom_hline(yintercept = benchmark_scurve,
     linetype="solid", 
     color = "black", linewidth=0.5, alpha = 0.5) +
  ylab(expression(d^{(2)})) +
  ggtitle("(d)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        plot.title = element_text(size = 12))
```

@@

@@

## A curious difference between tSNE, UMAP and PaCMAP revealer

@@

The tSNE layout, model in \gD{}, and a view of the fit in projections from $4\text{-}D$, for the five Gaussian cluster data. The model fits the separation and tries to *filled out* the clusters. (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/RASEE7N5MbM>). 
:::

@@

The UMAP layout, model in \gD{}, and a view of the fit in projections from $4\text{-}D$, for the five Gaussian cluster data. The model fits the separation and tries to *filled out* the clusters, but not as much as tSNE. (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/iG4bCPkJilw>). 
:::

@@

The PaCMAP layout, model in \gD{}, and a view of the fit in projections from $4\text{-}D$, for the five Gaussian cluster data. The model fits the separation and shows *flat* shaped clusters. (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/z07cKXi8EJQ>). 
:::

### 2024-07-22 paper.tex
Resize Figure 8

The benchmark value for removing low-density hexagons ranges between
\(0\) and \(1\). When analyzing how these benchmark values influence
model performance, it's essential to observe the change in MSE as the
benchmark value increases (see \textbf{?@fig-mse-scurve-lwd}). The MSE
shows a gradual increase as the benchmark value progresses from \(0\) to
\(1\). Evaluating this rate of increase is important. If the increment
is not considerable, the decision might lean towards retaining
low-density hexagons.

Furthermore, selecting the benchmark value for removing low-density
hexagons is important. Removing unnecessary bins may lead to the
formation of long edges and an uneven \gD{} model. Hence, rather than
solely relying on the benchmark value to identify hexagons for removal,
it's essential to consider the standard number of points in the
neighboring hexagons of the identified low-density bins (see
\textbf{?@fig-lwd-scurve} (b)). If neighboring bins also show low
counts, only those bins will be removed. The remaining bins are used to
construct the \gD{} model.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Plot the distribution of the 2D Euclidean distances
  (Figure~\ref{fig-param-scurve} (d)).
\item
  Identify the first largest difference between consecutive distance
  values.
\item
  Take the distance value corresponding to this difference as the
  benchmark value.
\end{enumerate}

@@

\section{A curious difference between t-SNE, UMAP and PaCMAP
revealer}\label{a-curious-difference-between-t-sne-umap-and-pacmap-revealer}

@@

\caption{\label{fig-gau-tsne-sc}The tSNE layout, model in \gD{}, and a
view of the fit in projections from \(4\text{-}D\), for the five
Gaussian cluster data. The model fits the separation and tries to
\emph{filled out} the clusters. (The \textbf{langevitour} software is
used to view the data with a tour, and the full video is available at
(\url{https://youtu.be/RASEE7N5MbM}).}

@@

\caption{\label{fig-gau-umap-sc}The UMAP layout, model in \gD{}, and a
view of the fit in projections from \(4\text{-}D\), for the five
Gaussian cluster data. The model fits the separation and tries to
\emph{filled out} the clusters, but not as much as tSNE. (The
\textbf{langevitour} software is used to view the data with a tour, and
the full video is available at (\url{https://youtu.be/iG4bCPkJilw}).}

@@

\caption{\label{fig-gau-pacmap-sc}The PaCMAP layout, model in \gD{}, and
a view of the fit in projections from \(4\text{-}D\), for the five
Gaussian cluster data. The model fits the separation and shows
\emph{flat} shaped clusters. (The \textbf{langevitour} software is used
to view the data with a tour, and the full video is available at
(\url{https://youtu.be/z07cKXi8EJQ}).}

>>>>

The benchmark value for removing low-density hexagons ranges between
\(0\) and \(1\). When analyzing how these benchmark values influence
model performance, it's essential to observe the change in MSE as the
benchmark value increases (Figure~\ref{fig-param-scurve}). The MSE shows
a gradual decrease as the benchmark value goes from \(1\) to \(0\).
Evaluating this rate of increase is important. If the increment is not
considerable, the decision might lean towards retaining low-density
hexagons.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Plot the distribution of the \gD{} Euclidean distances
  (Figure~\ref{fig-param-scurve} (d)).
\item
  Identify the first largest difference between consecutive distance
  values.
\item
  Take the distance value corresponding to this difference as the
  benchmark value.
\end{enumerate}

@@

\section{A curious difference between tSNE, UMAP and PaCMAP
revealer}\label{a-curious-difference-between-tsne-umap-and-pacmap-revealer}

@@

\caption{\label{fig-gau-tsne-sc}The tSNE layout, model in \gD{}, and a
view of the fit in projections from \(4\text{-}D\), for the five
Gaussian cluster data. The model fits the separation and tries to
\emph{filled out} the clusters. (The \textbf{langevitour} software is
used to view the data with a tour, and the full video is available at
\url{https://youtu.be/RASEE7N5MbM}).}

@@

\caption{\label{fig-gau-umap-sc}The UMAP layout, model in \gD{}, and a
view of the fit in projections from \(4\text{-}D\), for the five
Gaussian cluster data. The model fits the separation and tries to
\emph{filled out} the clusters, but not as much as tSNE. (The
\textbf{langevitour} software is used to view the data with a tour, and
the full video is available at \url{https://youtu.be/iG4bCPkJilw}).}

@@

\caption{\label{fig-gau-pacmap-sc}The PaCMAP layout, model in \gD{}, and
a view of the fit in projections from \(4\text{-}D\), for the five
Gaussian cluster data. The model fits the separation and shows
\emph{flat} shaped clusters. (The \textbf{langevitour} software is used
to view the data with a tour, and the full video is available at
\url{https://youtu.be/z07cKXi8EJQ}).}
